---
title: "Week13"
author: "Candice Lei (A18585167)"
date: "2026-05-14"
format: pdf
toc: true
---

## Import countData and colData

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names = 1)
metadata <- read.csv("airway_metadata.csv")

head(counts)

head(metadata)
```

> Q1. How many genes are in this dataset?

There are 38694 genes in this dataset.

```{r}
dim(counts)
nrow(counts)
```

> Q2. How many ‘control’ cell lines do we have?

There are 4 'control' cell lines.

```{r}
head(metadata)
sum(metadata$dex == "control")
```

## Toy Differential Gene Expression

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

```{r}
library(dplyr)
control <- metadata %>% filter(dex=="control")
control.counts <- counts %>% select(control$id) 
control.mean <- rowSums(control.counts)/4
head(control.mean)
```

> Q3. How would you make the above code in either approach more robust? Is there a function that could help here?

I prefer the dplyr version because it is easier to read. A function that could help is `rowMeans()`. This function automatically calculates the mean across the selected columns.

```{r}
library(dplyr)

control <- metadata %>% filter(dex == "control")
control.counts <- counts %>% select(all_of(control$id))
control.mean <- rowMeans(control.counts)

head(control.mean)
```

> Q4. Follow the same procedure for the `treated` samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called `treated.mean`)

```{r}
treated <- metadata[metadata[, "dex"] == "treated", ]
treated.counts <- counts[, treated$id]
treated.mean <- rowMeans(treated.counts)

head(treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

> Q5a. Create a scatter plot showing the mean of the treated samples against the mean of the control samples.

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

> Q5b. You could also use the **ggplot2** package to make this figure producing the plot below. What **geom\_?()** function would you use for this plot?

I would use `geom_point()` for this plot.

```{r}
library(ggplot2)

ggplot(meancounts, aes(x = control.mean, y = treated.mean)) + geom_point()
```

> Q6. Try plotting both axes on a log scale. What is the argument to **plot()** that allows you to do this?

The argument is `log = "xy"`. This puts both the x-axis and y-axis on a log scale.

```{r}
plot(meancounts[,1],
     meancounts[,2],
     xlab = "log Control",
     ylab = "log Treated",
     log = "xy")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

> Q7. What is the purpose of the `arr.ind` argument in the **which()** function call above? Why would we then take the first column of the output and need to call the **unique()** function?

The purpose of the `arr.ind()` argument is to tell R to return the row and column locations of 0 values. We need to take the first column of the output because it contains the row numbers of genes with 0 expression. We then call the `unique()` function so each gene row is removed once.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```

> Q8. Using the `up.ind` vector above can you determine how many up regulated genes we have at the greater than 2 fc level?

There are 250 up regulated genes.

```{r}
sum(up.ind)
```

> Q9. Using the `down.ind` vector above can you determine how many down regulated genes we have at the greater than 2 fc level?

There are 367 down regulated genes.

```{r}
sum(down.ind)
```

> Q10. Do you trust these results? Why or why not?

I would not trust these results because we have only been using fold change. Fold change tells us how large the difference is between treated and control samples, but it doesn't tell us if the difference is statistically significant. A gene can have a large log2 fold change, such as greater than 2 or less than -2, but that does not mean the change is statistically significant. We have not calculated p-values or adjusted p-values yet, so these results could be misleading. A better analysis would include statistical testing, like DESeq2, to calculate adjusted p-values.

## Setting Up for DESeq

```{r}
library(DESeq2)
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```

## Principal Component Analysis (PCA

```{r}
vsd <- vst(dds, blind = FALSE)
plotPCA(vsd, intgroup = c("dex"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("dex"), returnData=TRUE)
head(pcaData)
```

```{r}
# Calculate percent variance per PC for the plot axis labels
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData) +
  aes(x = PC1, y = PC2, color = dex) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

## DESeq Analysis

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res.df <- as.data.frame(res)
head(res.df)

summary(res)
```

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

## Adding Annotation Data

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
keys=row.names(res), # Our genenames
keytype="ENSEMBL",        # The format of our genenames
column="SYMBOL",          # The new format we want to add
multiVals="first")

head(res)
```

> Q11. Run the **mapIds()** function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and `res$genename`.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,keys=row.names(res),column="ENTREZID", keytype="ENSEMBL",
multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,keys=row.names(res),column="UNIPROT",keytype="ENSEMBL",
multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,keys=row.names(res),column="GENENAME",keytype="ENSEMBL",
multiVals="first")

head(res)
```

```{r}
ord <- order( res$padj )
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

## Data Visualization

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
library(EnhancedVolcano)
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    labSize = 3,
    pointSize = 1.5)
```
