---
title: 'Lab09: Candy Mini Project'
author: "Candice Lei (A18585167)"
date: "2026-04-30"
format: pdf
toc: true
---

## Import Candy Data

```{r}
candy_file <- "candy-data.csv"
candy <- read.csv(candy_file, row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

There are 85 candy types. 

```{r}
# Counts how many rows there are
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

There are 38 fruity candy types. 

```{r}
# Counts how many candy have fruit = 1
sum(candy$fruity)
```

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |>
  select(winpercent)
```
> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

My favorite candy is Snickers and it's winpercent value is 76.67378.

```{r}
candy["Snickers", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

The winpercent value for "Kit Kat" is 76.7686.

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The winpercent value for Tootsie Roll is 49.6535.

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent column looks to be on a different scale compared to the other columns. The majority of the other columns in the dataset have values ranging from 0 to 1 and winpercent ranges from 20 to 85. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A 0 represents the candy is not a chocolate and a 1 represents that the candy is chocolate/has chocolate for the candy$chocolate column.

## Exploratory Analysis

> Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(binwidth = 5, color = "white", fill = "lightblue") + 
  labs(title = "Histogram of Candy Win Percentage",
       x = "Win Percentage",
       y = "Count")
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values does not look perfectly symmetrical. 

> Q10. Is the center of the distribution above or below 50%?

Based on the `skim()` output, the median for p50 for winpercent is 47.83. The center of the distribution is below 50%. 

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[as.logical((candy$chocolate))])
mean(candy$winpercent[as.logical(candy$fruity)])
```

On average chocolate candy is ranked higher than fruit candy. This is because chocolate candy have a higher mean winpercent, 60.92, compared to the winpercent for fruit candy, 44.12.

> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

The difference is statistically significant. The p-value is less than 0.05, which means chocolate candy have a significantly higher average winpercent then fruit candy. 

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

The least 5 liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

```{r}
library(dplyr)

candy |>
  arrange(winpercent) |>
  head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |>
  arrange(desc(winpercent)) |>
  head(5)
```
The top 5 favorite candy are Reese's Peanut Butter Cup, Reese's Miniatures, Twix, Kit Kat, and Snickers.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x = winpercent, y = reorder(rownames(candy), winpercent)) +
  geom_col() +
  labs(title = "Candy Rankings by Win Percentage",
       x = "Win Percentage",
       y = "Candy Type")
```

## Time to add some useful color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Boston Baked Beans.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst. 

## Taking a Look at PricePercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(pricepercent, winpercent, label = rownames(candy)) +
  geom_point(col = my_cols) +
  geom_text_repel(col = my_cols, size = 3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures has a very high winpercent and a low pricepercent. 

```{r}
candy[order(candy$pricepercent, -candy$winpercent), c("pricepercent", "winpercent")]
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate. Of those candies, Nik L Nip is the least popular, since it has the lowest winpercent.

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c("pricepercent", "winpercent")], n = 5)
```

## Exploring the Correlation Structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity are anti-correlated. 

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar are the most positively correlated. 

## Principal Component Analysis (PCA)

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$rotation[, 1])
plot(pca$x[, 1:2], col = my_cols, pch = 16)
```

```{r}
my_data <- cbind(candy, pca$x[, 1:3])
p <- ggplot(my_data) +
  aes(x = PC1, y = PC2,
      size = winpercent / 100,
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)
p
```

```{r}
library(ggrepel)
p + geom_text_repel(size = 3.3, col = my_cols, max.overlaps = 7) +
  theme(legend.position = "none") +
  labs(title = "Halloween Candy PCA Space",
       subtitle = "Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)", 
       caption = "Data from 538")
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The original variables that were picked up strongly by PC1 in the positive direction are fruity, pluribus, and hard. These do make sense to me because fruity candy most of the time don't have chocolate and they come in multiple pieces or harder candy. I saw this relationship highlighted in the previous correlation plot where fruity candy was negatively correlated with chocolate. 

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col() +
  labs(title = "PCA Loadings for PC1",
       x = "PC1",
       y = "Variable") +
  theme(plot.title = element_text(hjust=0.5))
```

## Summary 
> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

The combination of characteristics that make a "winning" candy appear to be chocolate, in a bar form, and not fruity/hard. The chocolate candies have a higher winpercent compared to fruity candies. The t-test also showed the difference was statistically significant. The top ranked candies were also chocolate candy, like Reese's, KitKat, and Snickers with filling. The plots were able to help us visualize the ranking of candies and the comparison of winpercent to other variables, like pricepercent. The correlation plot showed that chocolate/winpercent are positively related and fruit/chocolate are negatively related. The PCA separated the chocolate/bar candies from the fruity/pluribus/hard candies along PC1. All of these different analyses support the conclusion that chocolate candies with filling appear to make a "winning" candy. 