---
title: 'Week 6: R functions'
author: "Candice Lei (PID: A18585167)"
date: "2026-04-16"
format: pdf
toc: true 
---

## Background

All functions in R have at least 3 things: - a *name* (we pick that and
use it to call the function) - input *argument* (one or more comma
separated inputs that go inside the brackets when we call the
function) - the *body* (the lines of R code that do the work of the
function)

## Write your first R function: add()

Here we will create a function to add some numbers. Let's call it
'add()'

Input arguments can be either **required** or **optional**. The later
have fall back default values that will be used if the user does not
specify them.

> Q1a. Your first version of the function should add two input numbers
> together. For example, add(4,7) should return 11.

```{r}
add <- function(x, y = 1000) {
  x + y
}
```

Can we use our new function:

```{r}
add(10,1000)
```

```{r}
add(10)
```

```{r}
add(4,7)
```

> Q1b. For you second version, adapt your first function so it can take
> a single input vector or two inputs as before. For example, add(4, 7)
> and add( c(4,7,10) ). [1 pt]

```{r}
add <- function(x, y = 0) {
  sum(x,y) 
  }
```

```{r}
add(4,7)
add(c(4,7,10))
```

> Q1c. Finally, on your own (outside of class) create a third version of
> your function that can add any number of inputs that the user
> provides. For example, add(1, 2, 3, -4) should return 2. Hint: explore
> the... (dots) argument or the base R function sum(). [2 pts]

```{r}
add <- function(...) {
  sum(...)
}
```

```{r}
add(1,2,3,-4)
```

We can explicitly set a **return** value output from a function (rather
than the default last line) by using the 'return()' function call.

```{r}
add <- function(...) {
  return(sum(x,y,...))
  cat("Is it break time yet?\n")
}
```

## Write a generate_dna() function

A useful function here is the "base R" 'sample()' function:

```{r}
sample(1:5, size=3)
```

```{r}
sample(1:5, size=60, replace = TRUE)
```

We can use this to make a random nucleotide sequence if we draw from
"A", "C", "G", and "T"...

```{r}
sample(x=c("A","C","G","T"), size = 10, replace = TRUE)
```

> Q2a. Your first version should return a multi-element vector of single
> character nucleotides. For example generate_dna(6) might return "A",
> "T", "T", "G", "A", "C". [1 pt]

```{r}
generate_dna <- function(len) {
  sample(x=c("A","C","G","T"), size = len, replace = TRUE)
}
```

```{r}
generate_dna(6)
```

> Q2b. Your second version should **optionally** be able to return
> either a multi-element vector of single character nucleotides (as
> before) or a **single character string** (not a vector of single
> letters but a singe vector of multiple letters). For example
> "AAGGCTG". [1 pt]

Functions that could be useful here are 'paste()', 'if()', 'cat()', and
'return()'

```{r}
generate_dna <- function(len=10) {
  sample(x=c("A","C","G","T"), size=len, replace = TRUE)
}
```

```{r}
generate_dna <- function(len=10, single.element=TRUE) {
  ans <- sample(x=c("A","C","G","T"), size=len, replace = TRUE)
  # cat("Hello....")
  if(single.element) {
    # cat("is it me you are looking for...")
    ans <- paste (ans, collapse = "")
  }
  return(ans)
}
```

```{r}
generate_dna(single.element = FALSE)
```

```{r}
paste(c("A", "C", "G"), collapse = "---")
```

> Q2c. Finally, create a final version of your function that prints out
> a FASTA format sequence with an id line indicating the sequence
> length. For example: \>len9 CGAAGGCTG Be sure to include clear
> comments that explain each step of your function. [2 pts]

```         
>len9
CGAAGGCTG
```

```{r}
cat("hello \n there")
```

```{r}
generate_dna <- function(len=10, single.element=TRUE) {
  ans <- sample(x=c("A","C","G","T"), size=len, replace = TRUE)
  
  if(single.element) {
  
    ans <- paste (ans, collapse = "")
  }
  ## Format as FASTA with an ID line
  cat( paste(">len", len, "\n", sep = ""))
  cat(ans)
  cat("\n")
  ##
  
  return(ans)
}
```

```{r}
x <- generate_dna(20)
```

## Write a 'generate_protein()' function

> Q3. Write a function 'generate_protein()' that returns a random
> peptide/protein sequence of a length specified by the user. For
> example 'generate_protein(6)' might return '"WQRTAG"'. Your function
> should:

```{r}
generate_protein <- function(len=9) {
  aa <- c("A", "R", "N", "D", "C", "E", "Q", "G", "H", "I", "L", "K", "M", "F", "P", "S", "T", "W", "Y", "V")
  
  ans <- sample(x=aa, size=len, replace = TRUE)
  paste(ans, collapse = "")
}
```

```{r}
generate_protein(5)
```

## Generate random protein sequences of length 6 to 13

> Q4.Adapt and use your 'generate_protein()' function to generate a
> series of random protein sequences ranging from 6 to 13 amino acids in
> length (one sequence per length). Take advantage of the base R
> function 'for()' or 'sapply()' so that you do not have to call
> 'generate_protein()' eight times by hand.

```{r}
for(l in 6:13) {
  cat(">",l, "\n", sep="")
  cat(generate_protein(l), "\n")
}
```

```{r}
generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
generate_protein(13)
```

## BLASTp search against nr — are your peptides “unique in nature”?

> | Length (aa) | Best hit % identity | Best hit % coverage | Unique? Y/N |
> |-------------|---------------------|---------------------|-------------|
> | 6           | 100%                | 100%                | N           |
> | 7           | 100%                | 100%                | N           |
> | 8           | 87.50%              | 100%                | Y           |
> | 9           | 88.89%              | 100%                | Y           |
> | 10          | 100%                | 90%                 | Y           |
> | 11          | 90%                 | 91%                 | Y           |
> | 12          | 90%                 | 83%                 | Y           |
> | 13          | 100%                | 77%                 | Y           |

> Q5a. At which sequence length do your randomly generated peptides
> start to look “unique in nature” (i.e. no 100% coverage + 100%
> identity hit)? [1 pt]

At sequence length 8, the peptides start to look "unique in nature".

> Q5b. Speculate why very short random peptides are almost always found
> in nr, while longer ones typically are not. Your answer should refer
> both to the size of the sequence space (20𝐿 for a peptide of length 𝐿) and to the size of the known protein universe. [1 pt]

Very short random peptides are almost always found in nr because their sequence space (20 L) is still small enough that the known protein universe has already sampled a lot of those possibilities by chance. For the longer peptide, it's not typically always found in nr because it grows exponentially by 20L. This makes it larger than the total number of sequences in known proteins. 

## Connecting your findings to immunology

> Q6. In 3–6 sentences total and using your Q5 data and the reasoning
> from Q5b, what do you think this minimum length is and why might it be a bad design choice for the immune system to present very short peptides? [3 pt]

I think the minimum length is around 7-8 amino acids because anything below that length, the random peptides are too common in the known protein universe that they aren't distinctive. It could be a bad design choice for the immune system to present very short peptides because the immune system needs peptides that are distinctive to be able to distinguish self from pathogens. If the presented peptides are too short, then there wouldn't be enough information and the chance of false matches would increase.
