---
title: "Class 08: Breast Cancer Mini Project"
author: "Candice Lei (A18585167)"
format: pdf
toc: true
---

## Background

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class.

Today we will analyze a biopsy data-set from fine needle aspiration (FNA) of a breast mass.

## Data Import

The data is made available as a CSV file for download. We can read this in using `read.csv()`:

```{r}
fna.data <- "WisconsinCancer.csv"
```

```{r}
wisc.df <- read.csv(fna.data, row.names = 1)
```

```{r}
head(wisc.df, 4)
```

Make sure we remove or exclude the `diagnosis` column from the data-set that we use for further analysis - this is the expert diagnosis as either M or B:

```{r}
wisc.data <- wisc.df[, -1]
diagnosis <- as.factor( wisc.df$diagnosis )
```

## Exploratory Data Analysis

> Q1. How many observations are in this dataset?

569 observations

```{r}
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

212 malignant diagnosis

```{r}
table(wisc.df$diagnosis)
sum( wisc.df$diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with `_mean`?

10 variables

We can use the `grep()` function to help us here:

```{r}
length( grep("_mean$", colnames(wisc.df), value = TRUE) )
```

## Principal Component Analysis (PCA)

We need to scale our data before PCA with the `scale=TRUE` argument to `prcomp()`

```{r}
colMeans(wisc.data)
apply(wisc.data, 2, sd)
```

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

PC1 captures about 44.3% of the original variance.

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs are required to describe at least 70% of the original variance.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs are required to describe at least 90% of the original variance.

## Interpreting PCA Results

Now you will use some visualizations to better understand your PCA model. A common visualization for PCA results is the so-called biplot.

Create a biplot of the `wisc.pr` using the `biplot()` function.

```{r}
biplot(wisc.pr)
```

Let's see the "PC score plot"

```{r}
library(ggplot2)
ggplot(wisc.pr$x) + 
  aes(PC1, PC2, col = diagnosis) +
  geom_point()
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The two clusters stood out to me on this plot. It is kind of difficult to understand because there are a lot of overlapping points in the plot, making it look crowded. It is also hard to know which original variables are driving the pattern.

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

I noticed that there is still separation between benign and malignant samples, along PC1. PC1 vs PC3 is less clear compared to PC1 vs PC2 because PC3 has less variance.

```{r}
ggplot(wisc.pr$x) + 
  aes(PC1, PC3, col = diagnosis) +
  geom_point()
```

## Variance Explained

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component
pve <- pr.var / sum(pr.var)

# Plot variance
plot(pve,
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     ylim = c(0,1),
     type = "o")

#Alternative scree plot of same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

## Communicating PCA Results

> Q9. For the first principal component, what is the component of the loading vector (i.e. `wisc.pr$rotation[,1]`) for the feature `concave.points_mean`? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

The loading for `concave.points_mean` is -0.2608538. Yes, there are features with larger contributions than this one, such as `concavity_mean`.

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

## Hierarchical Clustering

```{r}
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
plot(wisc.hclust)
```

## Results of Hierarchical Clustering

> Q10. Using the `plot()` and `abline()` functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h = 19, col = "red", lty = 2)
```

## Selecting Number of Clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Which method gives your favorite results for the same `data.dist` dataset? Explain your reasoning.

The ward.D2 method gives my favorite results for the dataset because it forms compact and well-separated clusters. It also minimizes the cluster variance.

```{r}
wisc.hclust.ward <- hclust(data.dist, method = "ward.D2")
plot(wisc.hclust.ward)
```

## Combining Methods

```{r}
wisc.pr.hclust <- hclust(
  dist(wisc.pr$x[, 1:min(which(summary(wisc.pr)$importance[3, ] >= 0.90))]),
  method = "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k = 2)
table(grps)
```

```{r}
table(grps, diagnosis)
```

```{r}
ggplot(wisc.pr$x) + 
  aes(PC1, PC2) + 
  geom_point(col = grps)

# Use distance along first 7PCs for clustering
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")

# Cut into 2 clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
```

> Q13. How well does the newly created `hclust` model with two clusters separate out the two “M” and “B” diagnoses?

It separates out the two "M" and "B" diagnoses pretty well

```{r}
table(cutree(wisc.pr.hclust, k = 2) , diagnosis)
```

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the `table()` function to compare the output of each model (`wisc.hclust.clusters` and `wisc.pr.hclust.clusters`) with the vector containing the actual diagnoses.

Compared to the PCA-based clustering, the hierarchical clustering models separate the the diagnoses okay. The PCA-based clustering shows a cleaner separation and separates the diagnoses better.

```{r}
table(wisc.hclust.clusters, diagnosis)
table(wisc.pr.hclust.clusters, diagnosis)
```

## Prediction

We will use the `predict()` function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
url <- "new_samples.csv"
new <- read.csv(url)
npc <-predict(wisc.pr, newdata = new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch = 16, cex = 3)
text(npc[,1], npc[,2], c(1,2), col = "white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

Based on my results, we should prioritize Patient 2 for follow up. Their clusters are closer to the malignant cancer samples compared to Patient 1.
