---
title: "Week5: Data Viz with ggplot2"
author: "Candice Lei (A18585167)"
format: pdf
toc: true
---

## Background

There are many graphic systems in R for making plots and figures. These include so-called base R graphics like the `plot()` function and add on packages like `ggplot2`
Let's compare how we make a simple figure with these two systems:
We can use the in-build 'cars' dataset:

```{r}
head(cars)
```
```{r}
plot(cars)
```
Before I can use ggplot2 I need to install it on my computer. To do this, we can use the function 'install.packages("ggplot2")
Once installed we need to load up the package into our R brain:

```{r}
library(ggplot2)
ggplot(cars)
```
Every ggplot has at least 3 layers:

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point()
```

## Add some custom features

Let's add a trend line that shows the relationship between speed and distance

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth()
```
> Q. Can you make the `geom_smooth()` function produce a linear straight line to fit the data and turn off the "gray" error region?

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(title = "Stopping Dist of Old Cars",
       x = "Speed (MPH)",
       y = "Distance (ft)")
```
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
```

```{r}
colnames(genes)
```

```{r}
ncol(genes)
```

```{r}
sum(genes$State == "up")
```
A useful new function in this context is the `table()` function:
```{r}
table(genes$State)
```
```{r}
round(table(genes$State)/nrow(genes) * 100, 2)
```
My first plot attempt
```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point()
```
```{r}
ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point() +
  scale_color_manual(values = c("purple", "gray", "orange")) +
  theme_bw() +
  labs(title = "Gene Expression Changes Upon Drug Treatment", x = "Control (No drug)", y = "Drug Treatment")
```
## Going Further

Here we read the famous gapminder dataset

```{r}
# File location online
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```
> Q. How many entries are in the dataset?

```{r}
nrow(gapminder)
```
> Q. How many different "country" entries are in this dataset?

```{r}
length(table(gapminder$country))
length(unique(gapminder$country))
```

Let's make our first plot of the entire dataset:
Plot of "gdpPercap" vs "LifeExp" colored by "continent"
```{r}
p <- ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
  geom_point(alpha = 0.3)
```

I can add more layers to 'p'

```{r}
p +
  facet_wrap(~continent)
```
Make a plot for 1977 and 2007 only
> Q. First use the `dplyr` package and the `filter()` function from that package to extract the year 2007.

```{r}
library(dplyr)

g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
g <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(g) +
  aes(gdpPercap, lifeExp, col = continent, size = pop) +
  geom_point(alpha = 0.3) +
  facet_wrap(~year)
```
> Q. Make a histogram of lifeExp colored by continent
> Q. Make a histogram of lifeExp faceted by continent

```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram()
```
```{r}
ggplot(gapminder) +
  aes(lifeExp, fill = continent) +
  geom_histogram()
```
```{r}
ggplot(gapminder) +
  aes(lifeExp) +
  geom_histogram() +
  facet_wrap(~continent)
```



